
---
title: "Simple Linear Regression with<br>Punta Cana Airbnb Data"
subtitle: "Analyzing the Relationship Between Bedrooms and Nightly Rate"
author: "David J. Gibbens<br>www.linkedin.com/in/djogibbens"
date: "04-12-2026"
output:
  ioslides_presentation:
    widescreen: true
    self_contained: false
# css: styles.css
---

<style type="text/css">
body p, div, h1, h2, h3, h4, h5 {
  color: black;
  font-family: "Modern Computer Roman";
}

slides > slide.title-slide hgroup h1 {
  color: #8C1D40;
}

h2 {
  color: #8C1D40;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE) # Suppresses code in the slide deck.

library(tidyverse) # Loads tidyverse which includes dplyr, ggplot2, readr, read_csv(), drop_na() and %>% pipe
library(plotly)    # Enables plotly which is a library that has interactive graphs

# Read Punta Cana dataset
pdc <- read_csv("PuntCanaListingsData-AirRoi.com-04-04-2026.csv")

# Code to Keep only the variables within the scope of this analysis for HW3 and establish a tibble

pdc_small <- pdc %>%     # Establishes the data frame (tibble) - table with rows and columns
  select(
    listing_id,          # Unique identifier
    bedrooms,            # X-variable (predictor for HW3)
    ttm_avg_rate,        # Y-variable (outcome)
    baths,               # Will be an X-variable in Project 1
    room_type            # Will be an X-variable in Project 1
  ) %>%                  # Pipe comes AFTER the closing parenthesis

  drop_na(bedrooms, ttm_avg_rate) # Remove missing values

# Quick sanity check: inspect the X and Y variables before modeling
# This confirms ranges, missing values, and basic distribution
summary(select(pdc_small, bedrooms, ttm_avg_rate))
```


## Introduction: Rationale for Completing This Analysis

Given the recent volatility in the stock market, I have decided to diversify my long‑term investment strategy by opening a self‑directed Roth IRA to purchase real estate in Punta Cana, Dominican Republic. My goal is to acquire a property that can operate as a profitable Airbnb rental while also appreciating in value over time.

To make informed investment decisions, I need to understand which property features — such as the number of bedrooms — make a listing more attractive to tourists and more lucrative as a short‑term rental. This analysis uses real Punta Cana listing data to explore how bedroom count relates to nightly rental rates, forming the foundation for more advanced modeling project.

## Data Overview
```{r}
summary(select(pdc_small, bedrooms, ttm_avg_rate))
```


## Scatterplot: Bedrooms vs Nightly Rate
```{r scatterplot, echo=FALSE, fig.width=7, fig.height=5}
ggplot(pdc_small, aes(x = bedrooms, y = ttm_avg_rate)) +
  geom_point(color = "steelblue", alpha = 0.7) +
  labs(
    title = "Bedrooms vs Nightly Rate",
    x = "Number of Bedrooms",
    y = "TTM Average Nightly Rate"
  ) +
  theme_minimal()
```

This scatterplot shows that listings with more bedrooms tend to command higher prices.


## Scatterplot with Regression Line
```{r regression_line, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=5}

ggplot(pdc_small, aes(x = bedrooms, y = ttm_avg_rate)) +
  geom_point(color = "steelblue", alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    title = "Bedrooms vs Nightly Rate with Regression Line",
    x = "Number of Bedrooms",
    y = "TTM Average Nightly Rate"
  ) +
  theme_minimal()
```

The regression line shows nightly rates increase as bedroom count rises.






## Regression Model Output
```{r}
# Fit simple linear regression model:
# Y = ttm_avg_rate (nightly rate)
# X = bedrooms (number of bedrooms)
# This model estimates how much nightly rate increases for each additional bedroom.
model <- lm(ttm_avg_rate ~ bedrooms, data = pdc_small)

# Display full regression output:
# - Coefficients (intercept and slope)
# - Standard errors
# - t-values and p-values (significance)
# - R-squared (variance explained)
# - F-statistic (overall model significance)
summary(model)
```

## Regression Equation

Using the coefficient estimates from the regression output:
- Intercept (β0) = -231.11
- Slope (β1) = 190.26

## The fitted regression equation is:



\[
\hat{Y} = -231.11 + 190.26 \cdot \text{Bedrooms}
\]

## Standard Error of the Regression

The **standard error of the regression (SER)** measures the typical size of the residuals — how far the observed nightly rates deviate from the regression line.



\[
SER = \sqrt{\frac{1}{n - 2} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2}
\]


The SER of 132.5 indicates that the model’s predictions typically deviate from actual nightly rates by about $132.
This is moderate relative to the overall price range, suggesting that bedroom count explains some—but not all—of the variation in nightly rates.
It means that we probably need to complete a multiple regression analysis.

## Interpretation of the Slope

The estimated slope for **bedrooms** is **190.26**.  
This means:

**For every additional bedroom, the predicted TTM average nightly rate increases by about \$190.26 on average.**

Because the p‑value for the slope is extremely small (\(< 2e-16\)), the relationship between bedrooms and nightly rate is **statistically significant**.



## Interpretation of the Intercept

The intercept is **–231.11**, which represents the predicted nightly rate when the number of bedrooms is zero.

This value is **not meaningful in a real‑estate context**, but it is required mathematically to anchor the regression line.



## Interpretation of R-Squared

The regression model has an R-squared value of **0.4077**.

This means:

**Approximately 40.8% of the variation in TTM average nightly rate is explained by differences in the number of bedrooms.**

This indicates a moderately strong relationship for a single‑predictor model.



## Interpretation of Overall Model Significance

The F-statistic is **129.4** with a p-value **< 2.2e-16**.

This indicates:

**The overall regression model is statistically significant**, meaning bedrooms meaningfully predict nightly rate.



## Conclusion

Based on this simple linear regression analysis:

- The number of bedrooms is a **statistically significant** predictor of TTM average nightly rate.  
- Each additional bedroom is associated with an estimated **\$190.26 increase** in nightly rate, on average.  
- The model explains approximately **40.8%** of the variation in nightly rates.  
- While bedrooms matter, other variables not included in this simple model also influence pricing.

Overall, there is a clear positive relationship between bedroom count and nightly rate, and this model provides a strong first step in understanding pricing dynamics in this market.

Next Steps are to perform a multiple regression analysis to identify all the key metrics which influence pricing for Airbnb's in Punta Cana.

## Why Include Diagnostic Plots?

To ensure that the simple linear regression model is appropriate for this analysis, it is important to evaluate whether the model assumptions are reasonably met.

- **Residual Plot** — checks whether the relationship is approximately linear and whether the residuals show constant variance across fitted values.  
- **Fitted vs Actual Plot** — visualizes how well the model’s predictions align with the true nightly rates.  
- **Purpose** — these diagnostics help confirm that the model is reliable and that the conclusions drawn from it are valid.

Including these plots strengthens the credibility of the analysis and provides a more complete understanding of model performance.


## Residual Plot
```{r residuals, echo=FALSE, fig.width=7, fig.height=5}
# Compute residuals and fitted values
residual_data <- data.frame(
  fitted = fitted(model),
  residuals = resid(model)
)

ggplot(residual_data, aes(x = fitted, y = residuals)) +
  geom_point(color = "steelblue", alpha = 0.7) +
  geom_hline(yintercept = 0, color = "darkred", linetype = "dashed") +
  labs(
    title = "Residuals vs Fitted Values",
    x = "Fitted Values (Predicted Nightly Rate)",
    y = "Residuals"
  ) +
  theme_minimal()
```

Residuals scatter evenly around the zero line, showing no clear pattern or bias.

## Fitted vs Actual Plot
```{r fitted_vs_actual, echo=FALSE, fig.width=7, fig.height=5}
# Compute fitted values and actual values
fitted_data <- data.frame(
  actual = pdc_small$ttm_avg_rate,
  fitted = fitted(model)
)

ggplot(fitted_data, aes(x = fitted, y = actual)) +
  geom_point(color = "steelblue", alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, color = "darkred", linetype = "dashed") +
  labs(
    title = "Fitted vs Actual Nightly Rates",
    x = "Fitted Values (Predicted Nightly Rate)",
    y = "Actual Nightly Rate"
  ) +
  theme_minimal()
```

Points that fall close to the dashed 45‑degree line indicate accurate predictions.<br>  








## Interactive Plotly Scatterplot
```{r, message=FALSE, warning=FALSE}
plot_ly(
  data = pdc_small,
  x = ~bedrooms,
  y = ~ttm_avg_rate,
  type = "scatter",
  mode = "markers",
  marker = list(color = "steelblue")
) %>%
  add_lines(
    x = pdc_small$bedrooms,
    y = fitted(lm(ttm_avg_rate ~ bedrooms, data = pdc_small)),
    line = list(color = "red")
  ) %>%
  layout(
    title = "Interactive Scatterplot with Regression Line",
    xaxis = list(title = "Number of Bedrooms"),
    yaxis = list(title = "TTM Average Nightly Rate")
  )
```

Residuals follow the diagonal closely, indicating the errors are approximately normally distributed.

## Thank You

Questions or comments are welcome.
